---
title: "CI Project - AML biomarker test"
output: html_document
author: "Arran Tattersall"
---

set directory
```{r setup, include=FALSE, }
setwd("/rds/general/user/at5023/home/CancerInformatics/Assignment/AML")
getwd()
```

import TCGA data to directory via bash
```{bash}
#data
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LAML.gistic.tsv.gz ##copy number variation by gene (GISTIC)
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LAML.methylation450.tsv.gz ## methylation (450) data
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LAML.mutect2_snv.tsv.gz ## single nucleotide variation data
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LAML.htseq_fpkm-uq.tsv.gz ## RNAseq data
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LAML.GDC_phenotype.tsv.gz ## clinical data
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-LAML.survival.tsv ## survival data

#mapping
wget https://gdc-hub.s3.us-east-1.amazonaws.com/download/illuminaMethyl450_hg38_GDC ## Human methylation data TCGA suggested


gunzip *.gz ## unzip
```

load required packages 
```{r}
library(survival)
```

import data as R readable
```{r}
cnv.data.1 <-
  read.table(
    "TCGA-LAML.gistic.tsv",
    sep = "\t",
    header = TRUE,
    row.names = 1
  ) ## cnv data by gene

meth.data <-
  read.table(
    "TCGA-LAML.methylation450.tsv",
    sep = "\t",
    header = TRUE,
    row.names = 1
  ) ## methylation data

snv.data <- 
  read.table(
    "TCGA-LAML.mutect2_snv.tsv", 
    sep = "\t", 
    header = TRUE,
    row.names = NULL
  ) ## snv data

rna.data <-
  read.table(
    "TCGA-LAML.htseq_fpkm-uq.tsv",
    sep = "\t",
    header = TRUE,
    row.names = 1
  )
rna.data <- as.matrix(rna.data) ## RNAseq data

clin.data <-
  read.table(
    "TCGA-LAML.GDC_phenotype.tsv",
    sep = "\t",
    header = TRUE,
    row.names = 1
  )
clin.data$leukemia_french_american_british_morphology_code[clin.data$leukemia_french_american_british_morphology_code == "M0 Undifferentiated"] <- "M0"
rownames(clin.data) <-
  gsub(rownames(clin.data), pattern = "-", replace = ".") ## clinical data


surv.data <- 
  read.table(
    "TCGA-LAML.survival.tsv",
    sep = "\t",
    header = TRUE,
    row.names = 1
  )
rownames(surv.data) <-
  gsub(rownames(surv.data), pattern = "-", replace = ".") ## survival data
```

with survival package find associations with patient outcomes
```{r}
os.time <- surv.data[colnames(rna.data), "OS.time"] ## length of time before outcome 

os.event <- as.numeric(surv.data[colnames(rna.data), "OS"]) ## patient outcome 

cesc.os <- Surv(os.time, os.event)
```

find genes most significantly associated to patient outcome by a univariate model
```{r}
results.univariate <- array(NA, c(nrow(rna.data), 4))

colnames(results.univariate) <- c("HR", 
                                  "LCI", 
                                  "UCI", 
                                  "PVAL")

rownames(results.univariate) <- rownames(rna.data)

results.univariate <- as.data.frame(results.univariate)


for (i in 1:nrow(rna.data))
  
{
  coxphmodel <- coxph(cesc.os ~ as.numeric(rna.data[i, ]))
  
  results.univariate$HR[i] <- summary(coxphmodel)$coef[1, 2]
  
  results.univariate$LCI[i] <- summary(coxphmodel)$conf.int[1, 3]
  
  results.univariate$UCI[i] <- summary(coxphmodel)$conf.int[1, 4]
  
  results.univariate$PVAL[i] <- summary(coxphmodel)$coef[1, 5]
}


results.univariate <- as.data.frame(results.univariate)

results.univariate$FDR <-
  p.adjust(results.univariate$PVAL, method = "fdr")

results.univariate <-
  results.univariate[order(results.univariate$FDR, decreasing = FALSE),]

results.univariate[1:20, ]

# ENSG00000240436.1 

```

find clinical details most significantly associated to patient outcome by a multivariate model
```{r}
clin.data <- clin.data[colnames(rna.data), ] ## removing patients from clin.data which are not present in rna.data

age <- as.numeric(clin.data$age_at_initial_pathologic_diagnosis) ## age

treatment.history <- as.factor(clin.data$history_of_neoadjuvant_treatment) ## previous treatment

lymphocyte.count <- as.numeric(clin.data$lab_procedure_abnormal_lymphocyte_result_percent_value) ## abnormal lymphocyte count

blast.count <- as.numeric(clin.data$lab_procedure_blast_cell_outcome_percentage_value) ## blast count

bm.band <- as.numeric(clin.data$lab_procedure_bone_marrow_band_cell_result_percent_value) ## marrow band

bm.basophil <- as.numeric(clin.data$lab_procedure_bone_marrow_basophil_result_percent_value) ## bone marrow basophil count

bm.blast <- as.numeric(clin.data$lab_procedure_bone_marrow_blast_cell_outcome_percent_value) ## bone marrow blast count

cellularity <- as.numeric(clin.data$lab_procedure_bone_marrow_cellularity_outcome_percent_value) ## % bone marrow cellularity

bm.eosinophil <- as.numeric(clin.data$lab_procedure_bone_marrow_lab_eosinophil_result_percent_value) ## marrow band

bm.lymphocyte <- as.numeric(clin.data$lab_procedure_bone_marrow_lymphocyte_outcome_percent_value) ## bone marrow lympocyte count

bm.neutrophil <- as.numeric(clin.data$lab_procedure_bone_marrow_neutrophil_result_percent_value) ## bone marrow neutrophil count

diff <-  as.factor(clin.data$leukemia_french_american_british_morphology_code) ## FAB morphology

m0 <- grep("M0", diff)
m1 <- grep("M1", diff)
m2 <- grep("M2", diff)
m3 <- grep("M3", diff)
m4 <- grep("M4", diff)
m5 <- grep("M5", diff)
m6 <- grep("M6", diff)
m7 <- grep("M7", diff)

diff.high <- rep(0, nrow(clin.data))
diff.high[c(m2, m3, m4, m5, m6, m7)] <- 1 ## high maturation FAB classification

diff.low <- rep(0, nrow(clin.data))
diff.low[c(m0, m1)] <- 1 # low maturation FAB classification

diff.t3 <- rep(0, nrow(clin.data))
diff.t3[c(m5, m6, m7)] <- 1 ## after seeing coef values from diff summary, made new variable including 3 highest coef values

summary(coxph(cesc.os ~ age))$coef
summary(coxph(cesc.os ~ treatment.history))$coef
summary(coxph(cesc.os ~ lymphocyte.count))$coef
summary(coxph(cesc.os ~ blast.count))$coef
summary(coxph(cesc.os ~ bm.band))$coef
summary(coxph(cesc.os ~ bm.basophil))$coef
summary(coxph(cesc.os ~ bm.blast))$coef
summary(coxph(cesc.os ~ cellularity))$coef
summary(coxph(cesc.os ~ bm.eosinophil))$coef
summary(coxph(cesc.os ~ bm.lymphocyte))$coef
summary(coxph(cesc.os ~ bm.neutrophil))$coef
summary(coxph(cesc.os ~ diff))$coef
summary(coxph(cesc.os ~ diff.high))$coef
summary(coxph(cesc.os ~ diff.low))$coef
summary(coxph(cesc.os ~ diff.t3))$coef


## NB diff.high and diff.low have *-1 coef and same p-value, suprised at low blast influence
```

low.diff prognosis calculation
```{r}

results.multivariate.1 <- array(NA, c(nrow(rna.data), 4))

colnames(results.multivariate.1) <- c("HR", "LCI", "UCI", "PVAL")
rownames(results.multivariate.1) <- rownames(rna.data)

results.multivariate.1 <- as.data.frame(results.multivariate.1) 

for (i in 1:nrow(rna.data))
  
{
  coxphmodel <- coxph(cesc.os ~ rna.data[i, ] + diff.low)
  
  results.multivariate.1$HR[i] <- summary(coxphmodel)$coef[1, 2]
  
  results.multivariate.1$LCI[i] <- summary(coxphmodel)$conf.int[1, 3]
  
  results.multivariate.1$UCI[i] <- summary(coxphmodel)$conf.int[1, 4]
  
  results.multivariate.1$PVAL[i] <- summary(coxphmodel)$coef[1, 5]
}


results.multivariate.1 <- as.data.frame(results.multivariate.1)

results.multivariate.1$FDR <-
  p.adjust(results.multivariate.1$PVAL, method = "fdr")


results.multivariate.1 <-
  results.multivariate.1[order(results.multivariate.1$FDR, decreasing = FALSE), ]

results.multivariate.1[1:20, ] ## ENSG00000240436.1	HR 0.7397948	LCI0.6720393	HCI0.8143815	FDR4.472962e-05 RP11
```

high.diff prognosis calculation
```{r}

results.multivariate.2 <- array(NA, c(nrow(rna.data), 4))

colnames(results.multivariate.2) <- c("HR", "LCI", "UCI", "PVAL")
rownames(results.multivariate.2) <- rownames(rna.data)

results.multivariate.2 <- as.data.frame(results.multivariate.2) 

for (i in 1:nrow(rna.data))
  
{
  coxphmodel <- coxph(cesc.os ~ rna.data[i, ] + diff.high)
  
  results.multivariate.2$HR[i] <- summary(coxphmodel)$coef[1, 2]
  
  results.multivariate.2$LCI[i] <- summary(coxphmodel)$conf.int[1, 3]
  
  results.multivariate.2$UCI[i] <- summary(coxphmodel)$conf.int[1, 4]
  
  results.multivariate.2$PVAL[i] <- summary(coxphmodel)$coef[1, 5]
}


results.multivariate.2 <- as.data.frame(results.multivariate.2)

results.multivariate.2$FDR <-
  p.adjust(results.multivariate.2$PVAL, method = "fdr")


results.multivariate.2 <-
  results.multivariate.2[order(results.multivariate.2$FDR, decreasing = FALSE), ]

results.multivariate.2[1:20, ] ## same results as low.diff
```

FAB classes with highest coef prognosis calculation (m7 only has 1 sample, m6 only 2)
```{r}

results.multivariate.3 <- array(NA, c(nrow(rna.data), 4))

colnames(results.multivariate.3) <- c("HR", "LCI", "UCI", "PVAL")
rownames(results.multivariate.3) <- rownames(rna.data)

results.multivariate.3 <- as.data.frame(results.multivariate.3) 

for (i in 1:nrow(rna.data))
  
{
  coxphmodel <- coxph(cesc.os ~ rna.data[i, ] + diff.t3)
  
  results.multivariate.3$HR[i] <- summary(coxphmodel)$coef[1, 2]
  
  results.multivariate.3$LCI[i] <- summary(coxphmodel)$conf.int[1, 3]
  
  results.multivariate.3$UCI[i] <- summary(coxphmodel)$conf.int[1, 4]
  
  results.multivariate.3$PVAL[i] <- summary(coxphmodel)$coef[1, 5]
}


results.multivariate.3 <- as.data.frame(results.multivariate.3)

results.multivariate.3$FDR <-
  p.adjust(results.multivariate.3$PVAL, method = "fdr")


results.multivariate.3 <-
  results.multivariate.3[order(results.multivariate.3$FDR, decreasing = FALSE), ]

results.multivariate.3[1:20, ]
```

All point to ENSG00000240436.1, RP11-95I19.1 ( in GRCH38 it is RPL35AP14)
RPL35AP14 does not have any associated methylation, copy number variations or single nucleotide variations in the data,
and RPL35 (ENSG00000136942) is a known biomarker DOI: 10.3389/fendo.2023.1161269 


all diff calculation
```{r}
results.multivariate.x <- array(NA, c(nrow(rna.data), 4))

colnames(results.multivariate.x) <- c("HR", "LCI", "UCI", "PVAL")
rownames(results.multivariate.x) <- rownames(rna.data)

results.multivariate.x <- as.data.frame(results.multivariate.x) 

for (i in 1:nrow(rna.data))
  
{
  coxphmodel <- coxph(cesc.os ~ rna.data[i, ] + diff)
  
  results.multivariate.x$HR[i] <- summary(coxphmodel)$coef[1, 2]
  
  results.multivariate.x$LCI[i] <- summary(coxphmodel)$conf.int[1, 3]
  
  results.multivariate.x$UCI[i] <- summary(coxphmodel)$conf.int[1, 4]
  
  results.multivariate.x$PVAL[i] <- summary(coxphmodel)$coef[1, 5]
}


results.multivariate.x <- as.data.frame(results.multivariate.x)

results.multivariate.x$FDR <-
  p.adjust(results.multivariate.x$PVAL, method = "fdr")


results.multivariate.x <-
  results.multivariate.x[order(results.multivariate.x$FDR, decreasing = FALSE), ]

results.multivariate.x[1:20, ]

```

stratify patients by each expression
```{r}
summary(rna.data["ENSG00000107175.9", ])

creb.high <-
  as.numeric(rna.data["ENSG00000107175.9", ] > median(rna.data["ENSG00000107175.9", ]))

## found info CREB3 https://www.genecards.org/cgi-bin/carddisp.pl?gene=CREB3
```

plotting Kaplan–Meier graph
```{r}
png(
  "CESC_OS_byCREB3.png",
  width = 6,
  height = 6,
  units = 'in',
  res = 300
)
plot(
  survfit(cesc.os ~ creb.high),
  col = c("black", "red"),
  lwd = 2,
  mark.time = TRUE,
  xlab = "Overall Survival time (days)",
  ylab = "Proportion Surviving",
  main = "CREB3 in CESC"
)
legend(
  "topright",
  legend = c("CREB3-high", "CREB3-low"),
  col = c("red", "black"),
  lwd = 2
)
dev.off()
```

testing reasons for expression difference: methylation, copy number variation, single nucleotide variation
selecting methylation markers
```{r}
## ENSG00000107175.9 CREB3
meth.annot.1 <-
  read.csv("illuminaMethyl450_hg38_GDC",
    sep = "\t",
    header = TRUE,
    row.names = 1) ## convert file to R readable

meth.creb <-
  rownames(meth.annot.1[which(meth.annot.1$chrom == "chr9" & meth.annot.1$gene == "CREB3,MIR6853,TLN1"), ]) ## multiple genes on site, 7  sites found
```

generating new data sets using only sample data found in meth.data file
```{r}
rna.data2 <-
  rna.data[, which(is.element(colnames(rna.data), colnames(meth.data)))]
rna.data2 <- as.matrix(rna.data2[, order(colnames(rna.data2))])

meth.data2 <-
  meth.data[, which(is.element(colnames(meth.data), colnames(rna.data2)))]
meth.data2 <- as.matrix(meth.data2[, order(colnames(meth.data2))])

meth.data.creb <- meth.data2[meth.creb, ]

surv.data2 <-
  surv.data[which(is.element(rownames(surv.data), colnames(rna.data2))), ]
surv.data2 <- as.data.frame(surv.data2[colnames(surv.data2), ])
```

methylation correlation test
```{r}
creb.high2 <-
  as.numeric(as.numeric(rna.data2["ENSG00000107175.9", ]) > median(as.numeric(rna.data2["ENSG00000107175.9", ])))

results.meth <- array(NA, c(nrow(meth.data.creb), 4))
rownames(results.meth) <- rownames(meth.data.creb)
colnames(results.meth) <-
  c("Cor.CREB3",
    "Cor.test.CREB3",
    "Mean.high.CREB3",
    "Mean.low.CREB3")

for (i in 1:nrow(meth.data.creb))
{
  results.meth[i, 1] <-
    cor.test(as.numeric(rna.data2["ENSG00000107175.9", ]), as.numeric(meth.data.creb[i, ]), use ="c")$est
  results.meth[i, 2] <-
    cor.test(as.numeric(rna.data2["ENSG00000107175.9", ]), as.numeric(meth.data.creb[i, ]), use ="c")$p.value
}

results.meth[, 3] <-
  apply(meth.data.creb[, which(creb.high2 == 1)], 1, mean, na.rm = TRUE)

results.meth[, 4] <-
  apply(meth.data.creb[, which(creb.high2 == 0)], 1, mean, na.rm = TRUE)

summary(results.meth)
```

plotting methylation results
```{r}
png(
  "CESC_CREB3_ExpVsMeth_overlay.png",
  width = 6,
  height = 6,
  units = 'in',
  res = 300
)

plot(
  as.numeric(meth.data.creb["cg00833393", ]), ## longest y-axis for blank plot
  as.numeric(rna.data2["ENSG00000107175.9", ]),
  xlab = "DNA Methylation",
  ylab = "CREB3 RNAseq",
  main = "CESC_CREB3_ExpVsMeth",
  pch = 16
)

site_colors <- c()

for (site in c(
  "cg00833393",
  "cg00910297",
  "cg05219900",
  "cg12763402",
  "cg13694973",
  "cg14123850",
  "cg14307667"
)) ## combined plots
{
  color <- palette()[length(site_colors) + 1] ## unique differentiated colors for each site
  
  points(
    as.numeric(meth.data.creb[site, ]),
    as.numeric(rna.data2["ENSG00000107175.9", ]),
    col = color,
    pch = 16
  )

  abline(lm((as.numeric(rna.data2["ENSG00000107175.9", ])) ~ as.numeric(meth.data.creb[site, ])), 
  col = color, 
  lty = 2)

  site_colors <- c(site_colors, color)
}

legend(
  "bottomright",
  legend = c(
    "cg00833393",
    "cg00910297",
    "cg05219900",
    "cg12763402",
    "cg13694973",
    "cg14123850",
    "cg14307667"
  ),
  col = site_colors,
  pch = 16,
  bty = "n",
  title = "DNA Methylation Site"
)

dev.off()
## figure appears to corroborate a lack of correlation between expression and methylation
```

selecting for copy number variations
```{r}
creb.cnv <- cnv.data.1["ENSG00000107175.9", ]

summary(creb.cnv)
range(creb.cnv) ## no copy number variations present
```

selecting single nucleotide variation
```{r}
creb.snv <- snv.data["ENSG00000107175.9", ]

summary(creb.snv)## empty, no SNV data found
``` 

genecards lists 23 enhancers/promoters for CREB3. Altered expression or activation of these enhancers/promoters could explain the downstream changes in levels of CREB3 expression. Also would explain why methylation seems to have no correlation and why no copy number variation, or single nucleotide variation influences were found

